---
title: "Full Meta-Analysis Pipeline"
author: "Kira"
date: "2026-04-06"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: flatly
---


```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = TRUE,
  fig.width = 10, fig.height = 7, dpi = 150
)
```

# Setup and Data Loading {.tabset}

## Load packages

```{r packages}
# Core
library(tidyverse)
library(geosphere)

# Mixed models & meta-analysis
library(lme4)
library(lmerTest)
library(MuMIn)
library(metafor)
library(boot)

# Spatial
library(spdep)
library(vegan)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

# Network
library(igraph)

# Plotting
library(ggspatial)
library(scales)
library(circlize)
library(pheatmap)
library(ggrepel)
library(cowplot)
library(scatterpie)
library(ggalluvial)
library(treemapify)

# Tables
library(kableExtra)

# Paths
library(here)

# Output
dir.create("outputs", showWarnings = FALSE)
```

## Load and validate data

```{r load-data}
# ── FST data ─────────────────────────────────────────────────────────────────
fst_raw <- read_csv(
  here("data", "fst.csv"),
  col_types = cols(
    Site_A = col_character(), Site_B = col_character(),
    Fst = col_double(), significant = col_character(),
    .default = col_skip()
  )
)

# ── Genetic site metadata ────────────────────────────────────────────────────
genetic_meta <- read_csv(
  here("data", "geneticMeta.csv"),
  col_types = cols(
    region = col_character(), studyID = col_character(),
    lat = col_character(), lon = col_character(),
    lat_spec = col_double(), lon_spec = col_double(),
    spawnMode = col_character(), species = col_character(),
    author = col_character(), year = col_double(),
    geneMarkers = col_character(), .default = col_character()
  )
)

# ── Study metadata (marker type, spawn type) ─────────────────────────────────
all_studies <- read_csv(here("data", "allStudies.csv"), col_types = cols(.default = col_character()))

# ── Hydrodynamic connectivity ────────────────────────────────────────────────
hydro_raw <- read_csv(here("data", "hydrodynamic.csv"), col_types = cols(.default = col_character()))

# ── Region centroids (authoritative lat/lon per region) ───────────────────────
region_centroids <- read_csv(here("data", "regionData.csv"),
  col_types = cols(region = col_character(), lat = col_double(), lon = col_double())) %>%
  filter(!is.na(region), !is.na(lat), !is.na(lon))

# ── Basemap ───────────────────────────────────────────────────────────────────
basemap <- ne_countries(
  scale = "medium",
  country = c("Australia", "Indonesia", "East Timor"),
  returnclass = "sf"
)

cat("=== DATA LOADED ===\n")
cat("FST rows:", nrow(fst_raw), "\n")
cat("Genetic metadata rows:", nrow(genetic_meta), "\n")
cat("Study metadata rows:", nrow(all_studies), "\n")
cat("Hydrodynamic rows:", nrow(hydro_raw), "\n")
```

# Data Cleaning {.tabset}

This chunk performs ALL data cleaning in one place. Every downstream analysis
uses the objects created here. No further cleaning happens later.

## Clean site metadata

```{r clean-metadata}
# Study-level spawn type from allStudies (fallback for sites missing it in geneticMeta)
spawn_lookup <- all_studies %>%
  filter(studyType == "genetic", valuesAvail == "y") %>%
  select(study = studyID, spawn_from_study = spawnType) %>%
  mutate(spawn_from_study = na_if(spawn_from_study, "na"))

# CRITICAL: strip whitespace from ALL region names across ALL datasets
site_metadata <- genetic_meta %>%
  filter(!is.na(studyID), studyID != "") %>%
  mutate(region = str_trim(region)) %>%
  select(studyID, region, lat, lon, lat_spec, lon_spec,
         spawnMode, species, author, year, geneMarkers) %>%
  distinct(studyID, .keep_all = TRUE) %>%
  mutate(
    # Coerce lat/lon from character safely
    lat = suppressWarnings(as.numeric(lat)),
    lon = suppressWarnings(as.numeric(lon)),
    # Prefer site-specific coords; fall back to region centroid
    lat_use = coalesce(lat_spec, lat),
    lon_use = coalesce(lon_spec, lon),
    # Extract genus for taxon-level analysis
    taxon = word(species, 1),
    # Extract study prefix for joining
    study = str_extract(studyID, "^[a-z]{2,4}\\d{2}")
  ) %>%
  left_join(spawn_lookup, by = "study") %>%
  mutate(
    # Use site-level spawnMode where available; fall back to study-level spawnType
    spawnMode = coalesce(spawnMode, spawn_from_study)
  ) %>%
  select(-spawn_from_study)

cat("Unique sites:", nrow(site_metadata), "\n")
cat("Unique regions:", n_distinct(site_metadata$region), "\n")
cat("Regions:", paste(sort(unique(site_metadata$region)), collapse = ", "), "\n")
```

## Build marker type lookup

```{r marker-lookup}
# Marker type from allStudies.csv
marker_lookup <- all_studies %>%
  filter(studyType == "genetic", valuesAvail == "y") %>%
  select(study = studyID, geneticMarker = markerType) %>%
  filter(!is.na(study), !is.na(geneticMarker), study != "", geneticMarker != "na") %>%
  mutate(
    marker_category = case_when(
      str_detect(geneticMarker, regex("SNP", ignore_case = TRUE)) &
        str_detect(geneticMarker, regex("mtDNA|microsatellite|RNA", ignore_case = TRUE)) ~ "Mixed",
      str_detect(geneticMarker, regex("microsatellite", ignore_case = TRUE)) &
        str_detect(geneticMarker, regex("mtDNA", ignore_case = TRUE)) ~ "Mixed",
      str_detect(geneticMarker, regex("SNP", ignore_case = TRUE)) ~ "SNP",
      str_detect(geneticMarker, regex("microsatellite", ignore_case = TRUE)) ~ "Microsatellite",
      str_detect(geneticMarker, regex("allozyme", ignore_case = TRUE)) ~ "Allozyme",
      str_detect(geneticMarker, regex("mtDNA", ignore_case = TRUE)) ~ "mtDNA",
      str_detect(geneticMarker, regex("NGS", ignore_case = TRUE)) ~ "SNP",
      TRUE ~ "Other"
    ),
    # Marker resolution classification for manuscript
    marker_resolution = case_when(
      marker_category %in% c("SNP", "Mixed") ~ "High",
      marker_category == "Microsatellite" ~ "Medium",
      marker_category %in% c("Allozyme", "mtDNA") ~ "Low",
      TRUE ~ NA_character_
    )
  ) %>%
  select(study, geneticMarker, marker_category, marker_resolution) %>%
  distinct(study, .keep_all = TRUE)

cat("Marker lookup entries:", nrow(marker_lookup), "\n")
print(marker_lookup %>% select(study, marker_category, marker_resolution))
```

## Clean FST data and join metadata

```{r clean-fst}
# Check for unmatched sites
sites_in_fst <- unique(c(fst_raw$Site_A, fst_raw$Site_B))
sites_in_meta <- site_metadata$studyID
unmatched <- setdiff(sites_in_fst, sites_in_meta)
cat("Unmatched sites (in FST but not metadata):", length(unmatched), "\n")
if (length(unmatched) > 0) cat("  ", paste(unmatched, collapse = ", "), "\n")

# Clean: remove NAs, unmatched sites, clamp negative Fst to 0
fst_clean <- fst_raw %>%
  filter(!is.na(Fst), !is.na(significant)) %>%
  filter(!Site_A %in% unmatched, !Site_B %in% unmatched) %>%
  mutate(Fst = pmax(Fst, 0))  # Clamp negatives to 0

cat("FST values after cleaning:", nrow(fst_clean), "\n")

# Join metadata for both sites
fst_full <- fst_clean %>%
  left_join(site_metadata, by = c("Site_A" = "studyID"), suffix = c("", "_A")) %>%
  rename(Region_A = region, Lat_A = lat_use, Lon_A = lon_use,
         Spawn_A = spawnMode, Species_A = species, Taxon_A = taxon,
         Author_A = author, Year_A = year, Study_A = study) %>%
  select(-lat, -lon, -lat_spec, -lon_spec, -geneMarkers) %>%
  left_join(site_metadata, by = c("Site_B" = "studyID"), suffix = c("", "_B")) %>%
  rename(Region_B = region, Lat_B = lat_use, Lon_B = lon_use,
         Spawn_B = spawnMode, Species_B = species, Taxon_B = taxon,
         Author_B = author, Year_B = year, Study_B = study) %>%
  select(-lat, -lon, -lat_spec, -lon_spec, -geneMarkers) %>%
  filter(!is.na(Region_A), !is.na(Region_B)) %>%
  mutate(
    study = Study_A,  # Both sites share same study prefix
    same_region = Region_A == Region_B,
    region_pair = paste(pmin(Region_A, Region_B), pmax(Region_A, Region_B), sep = " <-> ")
  )

# Restrict to studies with valuesAvail == "y" in allStudies
valid_studies <- all_studies %>%
  filter(valuesAvail == "y") %>%
  pull(studyID)

fst_full <- fst_full %>% filter(study %in% valid_studies)

# Join marker type
fst_full <- fst_full %>%
  left_join(marker_lookup %>% select(study, marker_category, marker_resolution),
            by = "study")

cat("FST values with valid regions:", nrow(fst_full), "\n")
cat("Studies:", paste(sort(unique(fst_full$study)), collapse = ", "), "\n")
cat("Marker categories:\n")
fst_full %>% count(marker_category, sort = TRUE) %>% print()
```

## Split between- and within-region

```{r split-data}
fst_between <- fst_full %>% filter(!same_region)
fst_within  <- fst_full %>% filter(same_region)

cat("Between-region comparisons:", nrow(fst_between), "\n")
cat("Within-region comparisons:", nrow(fst_within), "\n")
cat("Unique between-region pairs:", n_distinct(fst_between$region_pair), "\n")
cat("Unique regions in between-region data:",
    n_distinct(c(fst_between$Region_A, fst_between$Region_B)), "\n")
```

## Calculate geographic distances

```{r geo-distances}
# Between-region data with valid coordinates
fst_between <- fst_between %>%
  filter(!is.na(Lat_A), !is.na(Lon_A), !is.na(Lat_B), !is.na(Lon_B)) %>%
  rowwise() %>%
  mutate(geo_dist_km = distHaversine(c(Lon_A, Lat_A), c(Lon_B, Lat_B)) / 1000) %>%
  ungroup() %>%
  filter(geo_dist_km > 0)

cat("Between-region with valid distances:", nrow(fst_between), "\n")
cat("Distance range:", round(min(fst_between$geo_dist_km)), "–",
    round(max(fst_between$geo_dist_km)), "km\n")
```

## Aggregate to region-pair level

```{r region-aggregate}
region_connectivity <- fst_between %>%
  group_by(region_pair) %>%
  summarise(
    n_comparisons = n(),
    n_studies = n_distinct(study),
    mean_fst = mean(Fst, na.rm = TRUE),
    sd_fst = sd(Fst, na.rm = TRUE),
    min_fst = min(Fst, na.rm = TRUE),
    max_fst = max(Fst, na.rm = TRUE),
    n_significant = sum(significant == "y"),
    studies_list = paste(sort(unique(study)), collapse = "; "),
    .groups = "drop"
  ) %>%
  mutate(
    # Extract Region_A and Region_B from the already-sorted region_pair key
    # (avoids dplyr sequential-evaluation bug when computing both inside summarise)
    Region_A = str_split_fixed(region_pair, " <-> ", 2)[, 1],
    Region_B = str_split_fixed(region_pair, " <-> ", 2)[, 2],
    connectivity = case_when(
      mean_fst < 0.05 ~ "High",
      mean_fst < 0.15 ~ "Moderate",
      TRUE ~ "Low"
    ),
    connectivity = factor(connectivity, levels = c("Low", "Moderate", "High"))
  )

cat("=== REGION-PAIR CONNECTIVITY SUMMARY ===\n")
cat("Total region pairs:", nrow(region_connectivity), "\n\n")
region_connectivity %>% count(connectivity) %>% print()
cat("\nPercentages:\n")
region_connectivity %>%
  count(connectivity) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  print()
```

## Clean hydrodynamic data

```{r clean-hydro}
valid_hydro_studies <- all_studies %>%
  filter(studyType == "dispersal", valuesAvail == "y") %>%
  pull(studyID)

hydro_conn <- hydro_raw %>%
  filter(conn == "y", studyID %in% valid_hydro_studies) %>%
  mutate(
    region_A = str_trim(region_A),
    region_B = str_trim(region_B),
    region_pair = paste(pmin(region_A, region_B), pmax(region_A, region_B), sep = " <-> "),
    connect_str = str_trim(connect_str),
    direction = str_trim(direction),
    strength_num = case_when(
      connect_str == "high" ~ 3L,
      connect_str == "mod"  ~ 2L,
      connect_str == "low"  ~ 1L,
      TRUE ~ NA_integer_
    )
  ) %>%
  select(studyID, region_A, region_B, region_pair, direction, connect_str, strength_num)

cat("Hydrodynamic connections:", nrow(hydro_conn), "\n")
cat("Strength breakdown:\n")
hydro_conn %>% count(connect_str) %>% print()
cat("\nDirection breakdown:\n")
hydro_conn %>% count(direction) %>% print()

# Aggregate to unique region pairs (max strength = consensus)
hydro_summary <- hydro_conn %>%
  group_by(region_pair) %>%
  summarise(
    n_hydro_studies = n_distinct(studyID),
    hydro_consensus = names(which.max(table(connect_str))),
    hydro_strength_num = max(strength_num, na.rm = TRUE),
    directions = paste(unique(direction), collapse = "; "),
    .groups = "drop"
  )

cat("\nUnique hydro region pairs:", nrow(hydro_summary), "\n")
cat("Hydro regions covered:",
    n_distinct(c(hydro_conn$region_A, hydro_conn$region_B)), "\n")
```

## Save cleaned data checkpoint

```{r save-checkpoint}
write_csv(fst_between, "outputs/01_fst_between_clean.csv")
write_csv(fst_within, "outputs/01_fst_within_clean.csv")
write_csv(region_connectivity, "outputs/01_region_connectivity.csv")
write_csv(hydro_summary, "outputs/01_hydro_summary.csv")

cat("=== DATA CLEANING COMPLETE ===\n")
cat("All downstream analyses use fst_between (n =", nrow(fst_between), ")\n")
cat("Region pairs:", nrow(region_connectivity), "\n")
```

# Literature Review {.tabset}

## Summary

```{r litrev-clean}
all_studies <- all_studies %>%
  mutate(
    publishedYear = as.integer(publishedYear),
    numSites      = suppressWarnings(as.integer(numSites)),
    sampleSize    = suppressWarnings(as.integer(str_extract(sampleSize, "^[0-9]+"))),
    studyType     = str_trim(studyType),
    valuesAvail   = str_trim(valuesAvail),
    spawnType     = str_trim(spawnType),
    location      = str_replace_all(location, regex("Houtman Abrolhos", ignore_case = TRUE), "Abrolhos")
  ) %>%
  mutate(
    marker_category = case_when(
      studyType != "genetic"                                                        ~ NA_character_,
      str_detect(markerType, regex("SNP|NGS", ignore_case = TRUE)) &
        str_detect(markerType, regex("microsatellite|mtDNA", ignore_case = TRUE))  ~ "Mixed",
      str_detect(markerType, regex("microsatellite", ignore_case = TRUE)) &
        str_detect(markerType, regex("mtDNA", ignore_case = TRUE))                 ~ "Mixed",
      str_detect(markerType, regex("SNP|NGS",            ignore_case = TRUE))      ~ "SNP",
      str_detect(markerType, regex("microsatellite",     ignore_case = TRUE))      ~ "Microsatellite",
      str_detect(markerType, regex("allozyme",           ignore_case = TRUE))      ~ "Allozyme",
      str_detect(markerType, regex("mtDNA",              ignore_case = TRUE))      ~ "mtDNA",
      TRUE                                                                          ~ "Other"
    )
  )

litrev_genetic   <- all_studies %>% filter(studyType == "genetic")
litrev_dispersal <- all_studies %>% filter(studyType == "dispersal")

year_min <- min(all_studies$publishedYear, na.rm = TRUE)
year_max <- max(all_studies$publishedYear, na.rm = TRUE)

cat("=== LITERATURE REVIEW SUMMARY ===\n")
cat("Total studies:", nrow(all_studies), "\n")
cat("  Genetic:", nrow(litrev_genetic), "\n")
cat("  Dispersal:", nrow(litrev_dispersal), "\n")
cat("Included (valuesAvail == y):", sum(all_studies$valuesAvail == "y"), "\n")
cat("Excluded (valuesAvail == n):", sum(all_studies$valuesAvail == "n"), "\n\n")
cat("Exclusion reasons:\n")
all_studies %>% filter(valuesAvail == "n") %>% count(studyType, excluReas) %>% print()
cat("\nPublication year range:",
    min(all_studies$publishedYear, na.rm = TRUE), "–",
    max(all_studies$publishedYear, na.rm = TRUE), "\n")
cat("Genetic — n sites (median):", median(litrev_genetic$numSites, na.rm = TRUE), "\n")
cat("Genetic — sample size range:",
    min(litrev_genetic$sampleSize, na.rm = TRUE), "–",
    max(litrev_genetic$sampleSize, na.rm = TRUE), "\n")
```

## Figure A — Publication timeline

```{r litrev-figA, fig.width=7, fig.height=3.5}
bw_type_cols <- c("genetic" = "black", "dispersal" = "grey45")

# Use only years that have at least one study — removes the 1986–2002 gap naturally
timeline_data <- all_studies %>%
  filter(!is.na(publishedYear)) %>%
  mutate(
    studyType = factor(studyType, levels = c("genetic", "dispersal")),
    yr_fct    = factor(publishedYear)
  ) %>%
  count(yr_fct, studyType) %>%
  complete(yr_fct, studyType, fill = list(n = 0L))

p_A <- ggplot(timeline_data, aes(x = yr_fct, y = n, fill = studyType)) +
  geom_col(position = "stack", width = 0.8, colour = NA) +
  scale_fill_manual(values = bw_type_cols,
                    name   = "Study type",
                    labels = c("Genetic", "Dispersal model")) +
  scale_y_continuous(breaks = scales::breaks_pretty(),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Publication year", y = "Number of studies") +
  theme_bw(base_size = 11) +
  theme(
    panel.grid       = element_blank(),
    axis.title       = element_text(size = 11, face = "bold"),
    axis.text.x      = element_text(size = 9, angle = 45, hjust = 1),
    axis.text.y      = element_text(size = 10),
    legend.title     = element_text(size = 9, face = "bold"),
    legend.text      = element_text(size = 8),
    legend.position  = "right"
  )

print(p_A)
ggsave("outputs/Figure_litrev_A_timeline.pdf", p_A,
       width = 130, height = 80, units = "mm", dpi = 600, bg = "white")
```

## Figure B — Geographic coverage

```{r litrev-figB, fig.width=9, fig.height=5}
# One row per study × region
litrev_region_studies <- all_studies %>%
  select(studyID, studyType, location) %>%
  separate_rows(location, sep = ";\\s*") %>%
  mutate(location = str_trim(location)) %>%
  filter(!is.na(location), location != "", str_to_lower(location) != "na") %>%
  inner_join(
    region_centroids %>% mutate(location = region),
    by = "location"
  ) %>%
  select(studyID, studyType, region, lat, lon)

# Wide format; radius ∝ sqrt(total) so area scales with study count
litrev_pie_data <- litrev_region_studies %>%
  count(region, studyType, lat, lon) %>%
  pivot_wider(names_from = studyType, values_from = n, values_fill = 0) %>%
  mutate(
    total  = rowSums(across(any_of(c("genetic", "dispersal")))),
    radius = 0.4 + 1.1 * sqrt(total / max(total))
  )

pie_cols_present <- intersect(c("genetic", "dispersal"), names(litrev_pie_data))

cat("=== GEOGRAPHIC COVERAGE ===\n")
print(litrev_pie_data %>% select(region, any_of(c("genetic","dispersal")), total, radius) %>%
        arrange(desc(total)))

# Arrange regions in a grid, ordered north → south by latitude
ncols_pie  <- 5
max_r      <- max(litrev_pie_data$radius)
spacing    <- max_r * 2.8   # guarantee no overlap

litrev_pie_grid <- litrev_pie_data %>%
  arrange(desc(lat)) %>%
  mutate(
    idx   = row_number() - 1L,
    x_pos = (idx %% ncols_pie) * spacing,
    y_pos = -(idx %/% ncols_pie) * spacing
  )

p_B <- ggplot() +
  geom_scatterpie(
    data      = litrev_pie_grid,
    aes(x = x_pos, y = y_pos, group = region, r = radius),
    cols      = pie_cols_present,
    colour    = "grey15",
    linewidth = 0.35
  ) +
  geom_text(
    data = litrev_pie_grid,
    aes(x = x_pos, y = y_pos - radius - 0.18, label = region),
    size  = 2.6, hjust = 0.5, vjust = 1, lineheight = 0.85
  ) +
  scale_fill_manual(
    values = c("genetic" = "black", "dispersal" = "grey45"),
    name   = "Study type",
    labels = c("Genetic", "Dispersal model")
  ) +
  coord_equal() +
  theme_void(base_size = 11) +
  theme(
    legend.title    = element_text(size = 9, face = "bold"),
    legend.text     = element_text(size = 8),
    legend.position = "right"
  )

print(p_B)
ggsave("outputs/Figure_litrev_B_geo.pdf", p_B,
       width = 220, height = 130, units = "mm", dpi = 600, bg = "white")
```

## Figure C — Taxonomic distribution

```{r litrev-figC, fig.width=5, fig.height=3.5}
litrev_taxa_data <- litrev_genetic %>%
  filter(!is.na(taxa), str_to_lower(taxa) != "na") %>%
  separate_rows(taxa, sep = ";\\s*") %>%
  mutate(taxa = str_trim(taxa)) %>%
  filter(taxa != "") %>%
  count(taxa) %>%
  mutate(taxa = fct_reorder(taxa, n))

p_C <- ggplot(litrev_taxa_data, aes(x = n, y = taxa)) +
  geom_col(fill = "grey30", colour = "grey20", linewidth = 0.3, width = 0.65) +
  geom_text(aes(label = n), hjust = -0.25, size = 3.2) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = "Number of genetic studies", y = "Taxon") +
  theme_bw(base_size = 11) +
  theme(
    panel.grid   = element_blank(),
    axis.title   = element_text(size = 11, face = "bold"),
    axis.text    = element_text(size = 10)
  )

print(p_C)
ggsave("outputs/Figure_litrev_C_taxa.pdf", p_C,
       width = 110, height = 80, units = "mm", dpi = 600, bg = "white")
```

## Figure D — Markers over time

```{r litrev-figD, fig.width=7, fig.height=3.5}
marker_grey <- c(
  "SNP"            = "black",
  "Mixed"          = "grey28",
  "Microsatellite" = "grey48",
  "mtDNA"          = "grey62",
  "Allozyme"       = "grey72"
)

d_breaks <- seq(1984, 2029, by = 5)
d_labels <- paste0(head(d_breaks, -1), "–", tail(d_breaks, -1) - 1)
d_labels[length(d_labels)] <- "2024–2026"   # cap final bin at actual data end

figD_data <- litrev_genetic %>%
  filter(!is.na(marker_category), marker_category != "Other",
         !is.na(publishedYear)) %>%
  mutate(
    year_bin        = cut(publishedYear, breaks = d_breaks, right = FALSE,
                          labels = d_labels, include.lowest = TRUE),
    marker_category = factor(marker_category,
                             levels = c("SNP", "Mixed", "Microsatellite", "mtDNA", "Allozyme"))
  )

p_D <- ggplot(figD_data, aes(x = year_bin, fill = marker_category)) +
  geom_bar(position = "stack", width = 0.72, colour = NA) +
  scale_fill_manual(values = marker_grey, name = "Marker type") +
  scale_y_continuous(breaks = scales::breaks_width(1),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Publication year", y = "Number of genetic studies") +
  theme_bw(base_size = 11) +
  theme(
    panel.grid     = element_blank(),
    axis.title     = element_text(size = 11, face = "bold"),
    axis.text.x    = element_text(size = 9, angle = 45, hjust = 1),
    axis.text.y    = element_text(size = 10),
    legend.title   = element_text(size = 9, face = "bold"),
    legend.text    = element_text(size = 8),
    legend.position = "right"
  )

print(p_D)
ggsave("outputs/Figure_litrev_D_markers.pdf", p_D,
       width = 130, height = 80, units = "mm", dpi = 600, bg = "white")
```

## Figure E — Taxa by study type (three alternatives)

```{r litrev-figE-data}
# Shared data prep for all three alternatives
litrev_taxa_studytype <- all_studies %>%
  filter(!is.na(taxa), str_to_lower(str_trim(taxa)) != "na") %>%
  separate_rows(taxa, sep = ";\\s*") %>%
  mutate(taxa = str_trim(taxa)) %>%
  filter(taxa != "") %>%
  count(taxa, studyType) %>%
  mutate(
    taxa      = fct_reorder(taxa, n, .fun = sum),
    studyType = factor(studyType, levels = c("genetic", "dispersal"))
  )

study_grey <- c("genetic" = "grey20", "dispersal" = "grey70")
study_labels <- c("genetic" = "Genetic", "dispersal" = "Dispersal model")
base_theme <- theme_bw(base_size = 11) +
  theme(
    panel.grid   = element_blank(),
    axis.title   = element_text(size = 11, face = "bold"),
    axis.text    = element_text(size = 10),
    legend.title = element_text(size = 9, face = "bold"),
    legend.text  = element_text(size = 8),
    legend.position = "right"
  )
```

### E1 — Diverging bar

```{r litrev-figE1, fig.width=5, fig.height=4}
p_E1 <- ggplot(litrev_taxa_studytype,
               aes(x = taxa, y = n, fill = studyType)) +
  geom_col(position = "stack", width = 0.8, colour = NA) +
  scale_fill_manual(values = bw_type_cols, name = "Study type",
                    labels = study_labels) +
  scale_y_continuous(breaks = scales::breaks_width(2),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Taxon", y = "No. of studies") +
  theme_bw(base_size = 11) +
  theme(
    panel.grid       = element_blank(),
    panel.border     = element_blank(),
    axis.line        = element_line(colour = "black", linewidth = 0.4),
    axis.title       = element_text(size = 11, face = "bold"),
    axis.text.x      = element_text(size = 9, angle = 45, hjust = 1),
    axis.text.y      = element_text(size = 10),
    legend.title     = element_text(size = 9, face = "bold"),
    legend.text      = element_text(size = 8),
    legend.position  = "right"
  )

print(p_E1)
ggsave("outputs/Figure_litrev_E1_diverging.pdf", p_E1,
       width = 130, height = 80, units = "mm", dpi = 600, bg = "white")
```

### Combined figure — A + E1 + D

```{r litrev-figCombined, fig.width=15, fig.height=3.5}
# Shared open-axis theme applied consistently to all three panels
theme_open <- theme_bw(base_size = 11) +
  theme(
    panel.grid      = element_blank(),
    panel.border    = element_blank(),
    axis.line       = element_line(colour = "black", linewidth = 0.4),
    axis.title      = element_text(size = 11, face = "bold"),
    axis.text.x     = element_text(size = 9, angle = 45, hjust = 1),
    axis.text.y     = element_text(size = 10),
    legend.title    = element_text(size = 9, face = "bold"),
    legend.text     = element_text(size = 8),
    legend.position = "right"
  )

p_comb_A <- p_A  + theme_open
p_comb_B <- p_E1 + theme_open
p_comb_C <- p_D  + theme_open

# align_plots() equalises the actual panel dimensions so all axis lines align
aligned <- align_plots(p_comb_A, p_comb_B, p_comb_C, align = "hv", axis = "tblr")

p_combined <- plot_grid(
  plotlist       = aligned,
  nrow           = 2,
  labels         = c("a", "b", "c"),
  label_size     = 11,
  label_fontface = "bold"
)

print(p_combined)
ggsave("outputs/Figure_litrev_combined.pdf", p_combined,
       width = 260, height = 160, units = "mm", dpi = 600, bg = "white")
```

### E2 — Unit dot chart

```{r litrev-figE2, fig.width=6, fig.height=4}
litrev_units <- litrev_taxa_studytype %>%
  uncount(n) %>%
  arrange(taxa, studyType) %>%
  group_by(taxa) %>%
  mutate(dot_x = row_number()) %>%
  ungroup()

p_E2 <- ggplot(litrev_units, aes(x = dot_x, y = taxa, colour = studyType)) +
  geom_point(size = 4, alpha = 0.85) +
  scale_colour_manual(values = study_grey, name = "Study type", labels = study_labels) +
  scale_x_continuous(breaks = scales::breaks_pretty(), name = "Number of studies") +
  labs(y = "Taxon") +
  base_theme +
  theme(panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.3))

print(p_E2)
ggsave("outputs/Figure_litrev_E2_units.pdf", p_E2,
       width = 140, height = 100, units = "mm", dpi = 600, bg = "white")
```

### E3 — Treemap

```{r litrev-figE3, fig.width=6, fig.height=4.5}
library(treemapify)
litrev_treemap <- litrev_taxa_studytype %>%
  mutate(label = paste0(taxa, "\n(", n, ")"))

p_E3 <- ggplot(litrev_treemap,
               aes(area = n, fill = studyType, label = label, subgroup = taxa)) +
  geom_treemap(colour = "white", linewidth = 0.8) +
  geom_treemap_subgroup_border(colour = "grey20", linewidth = 1.5) +
  geom_treemap_text(colour = "white", place = "centre", size = 9,
                    fontface = "italic", reflow = TRUE) +
  scale_fill_manual(values = study_grey, name = "Study type", labels = study_labels) +
  base_theme +
  theme(legend.position = "right")

print(p_E3)
ggsave("outputs/Figure_litrev_E3_treemap.pdf", p_E3,
       width = 150, height = 110, units = "mm", dpi = 600, bg = "white")
```

## Study table

```{r litrev-table}
gen_table <- litrev_genetic %>%
  arrange(publishedYear) %>%
  transmute(
    `Study`    = studyID,
    `Year`     = publishedYear,
    `Species`  = species,
    `Marker`   = markerType,
    `Regions`  = location,
    `Sites`    = numSites,
    `n`        = sampleSize,
    `Included` = if_else(valuesAvail == "y", "Yes", "No")
  )

kbl(gen_table, caption = "Genetic connectivity studies reviewed.") %>%
  kable_classic(full_width = FALSE, html_font = "Arial", font_size = 10) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(which(gen_table$Included == "No"), color = "grey55")

disp_table <- litrev_dispersal %>%
  arrange(publishedYear) %>%
  transmute(
    `Study`      = studyID,
    `Year`       = publishedYear,
    `Model`      = modelType,
    `Resolution` = modelRes,
    `Season`     = seasonModel,
    `Regions`    = location,
    `Included`   = if_else(valuesAvail == "y", "Yes", "No")
  )

kbl(disp_table, caption = "Dispersal modelling studies reviewed.") %>%
  kable_classic(full_width = FALSE, html_font = "Arial", font_size = 10) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(which(disp_table$Included == "No"), color = "grey55")
```

# Isolation-by-Distance Analysis

## Simple linear model

```{r ibd-simple}
ibd_lm <- lm(Fst ~ log10(geo_dist_km), data = fst_between)
lm_summ <- summary(ibd_lm)

cat("=== SIMPLE LINEAR MODEL ===\n")
cat("Slope:", round(coef(ibd_lm)[2], 4), "\n")
cat("R² =", round(lm_summ$r.squared, 4), "\n")
cat("F-statistic:", round(lm_summ$fstatistic[1], 2), "\n")
cat("p-value:", format(lm_summ$coefficients[2, 4], digits = 3), "\n")
```

## Mixed-effects model (study random intercept)

```{r ibd-mixed}
ibd_mixed <- lmer(Fst ~ log10(geo_dist_km) + (1 | study), data = fst_between)

cat("=== MIXED-EFFECTS MODEL: Fst ~ log10(distance) + (1|study) ===\n")
print(summary(ibd_mixed))

# R-squared
r2_mixed <- r.squaredGLMM(ibd_mixed)
cat("\nR² marginal (fixed):", round(r2_mixed[1, "R2m"], 4), "\n")
cat("R² conditional (fixed + random):", round(r2_mixed[1, "R2c"], 4), "\n")

# Variance components and ICC
vc <- as.data.frame(VarCorr(ibd_mixed))
var_study <- vc$vcov[vc$grp == "study"]
var_resid <- vc$vcov[vc$grp == "Residual"]
icc <- var_study / (var_study + var_resid)
cat("ICC (study):", round(icc, 3), "\n")
cat("Study variance:", round(var_study, 6), "\n")
cat("Residual variance:", round(var_resid, 6), "\n")
```

## Bootstrap confidence intervals

```{r ibd-bootstrap}
set.seed(42)
n_boot <- 1000

boot_slope <- boot(
  data = fst_between,
  statistic = function(d, i) coef(lm(Fst ~ log10(geo_dist_km), data = d[i, ]))[2],
  R = n_boot
)
boot_r2 <- boot(
  data = fst_between,
  statistic = function(d, i) summary(lm(Fst ~ log10(geo_dist_km), data = d[i, ]))$r.squared,
  R = n_boot
)

ci_slope <- boot.ci(boot_slope, type = "bca", conf = 0.95)
ci_r2    <- boot.ci(boot_r2,    type = "bca", conf = 0.95)

cat("=== BOOTSTRAP RESULTS (1000 nonparametric case resamples, BCa CIs) ===\n")
cat("IBD slope:", round(boot_slope$t0, 4),
    "[95% CI:", round(ci_slope$bca[4], 4), "–", round(ci_slope$bca[5], 4), "]\n")
cat("IBD R²:", round(boot_r2$t0, 4),
    "[95% CI:", round(ci_r2$bca[4], 4), "–", round(ci_r2$bca[5], 4), "]\n")
```

## Spatial autocorrelation (Moran's I)

```{r morans-i}
ibd_spatial <- fst_between %>%
  mutate(
    predicted = predict(ibd_mixed),
    residual = Fst - predicted,
    mid_lon = (Lon_A + Lon_B) / 2,
    mid_lat = (Lat_A + Lat_B) / 2
  )

coords_mat <- as.matrix(ibd_spatial[, c("mid_lon", "mid_lat")])
knn5 <- knearneigh(coords_mat, k = 5)
nb5 <- knn2nb(knn5)
listw5 <- nb2listw(nb5, style = "W")
moran5 <- moran.test(ibd_spatial$residual, listw5)

cat("=== MORAN'S I (k = 5 nearest neighbours) ===\n")
cat("Moran's I:", round(moran5$estimate[1], 4), "\n")
cat("Expected:", round(moran5$estimate[2], 4), "\n")
cat("p-value:", format(moran5$p.value, digits = 4), "\n")
cat("Interpretation:", ifelse(moran5$p.value < 0.05,
    "Significant spatial autocorrelation in residuals", "No spatial autocorrelation"), "\n")

# Sensitivity: k = 3 and k = 7
for (k in c(3, 7)) {
  knn_k <- knearneigh(coords_mat, k = k)
  nb_k <- knn2nb(knn_k)
  listw_k <- nb2listw(nb_k, style = "W")
  moran_k <- moran.test(ibd_spatial$residual, listw_k)
  cat(sprintf("  k=%d: I=%.4f, p=%s\n", k, moran_k$estimate[1], format(moran_k$p.value, digits = 3)))
}
```

## Mantel test (alternative IBD test)

```{r mantel-test}
# Build site-level matrices
sites_df <- bind_rows(
  fst_between %>% select(site = Site_A, lat = Lat_A, lon = Lon_A),
  fst_between %>% select(site = Site_B, lat = Lat_B, lon = Lon_B)
) %>%
  distinct(site, .keep_all = TRUE) %>%
  filter(!is.na(lat), !is.na(lon))

site_coords <- as.matrix(sites_df[, c("lon", "lat")])
rownames(site_coords) <- sites_df$site

geo_dist_mat <- distm(site_coords, fun = distHaversine) / 1000
rownames(geo_dist_mat) <- colnames(geo_dist_mat) <- sites_df$site

fst_mat <- matrix(NA, nrow = nrow(sites_df), ncol = nrow(sites_df),
                  dimnames = list(sites_df$site, sites_df$site))
for (i in seq_len(nrow(fst_between))) {
  sa <- fst_between$Site_A[i]; sb <- fst_between$Site_B[i]
  if (sa %in% rownames(fst_mat) & sb %in% colnames(fst_mat)) {
    fst_mat[sa, sb] <- fst_between$Fst[i]
    fst_mat[sb, sa] <- fst_between$Fst[i]
  }
}
diag(fst_mat) <- 0

# Keep sites with >1 comparison
good_sites <- rownames(fst_mat)[rowSums(!is.na(fst_mat)) > 1]
fst_sub <- fst_mat[good_sites, good_sites]
geo_sub <- geo_dist_mat[good_sites, good_sites]

# Impute missing with mean (standard for incomplete matrices)
fst_sub_complete <- fst_sub
fst_sub_complete[is.na(fst_sub_complete)] <- mean(fst_sub, na.rm = TRUE)

set.seed(42)
mantel_raw <- mantel(as.dist(fst_sub_complete), as.dist(geo_sub),
                     method = "spearman", permutations = 9999)
mantel_log <- mantel(as.dist(fst_sub_complete), as.dist(log10(geo_sub + 1)),
                     method = "spearman", permutations = 9999)

cat("=== MANTEL TESTS (9999 permutations, Spearman) ===\n")
cat("Raw distance:  r =", round(mantel_raw$statistic, 4), " p =", mantel_raw$signif, "\n")
cat("Log distance:  r =", round(mantel_log$statistic, 4), " p =", mantel_log$signif, "\n")
```

## Identify barriers and corridors

```{r barriers-corridors}
ibd_spatial <- ibd_spatial %>%
  mutate(
    sd_resid = sd(residual),
    deviation_type = case_when(
      residual >  2.0 * sd_resid ~ "Strong barrier",
      residual >  1.5 * sd_resid ~ "Moderate barrier",
      residual < -2.0 * sd_resid ~ "Strong corridor",
      residual < -1.5 * sd_resid ~ "Moderate corridor",
      TRUE ~ "Expected (IBD)"
    )
  )

cat("=== DEVIATION SUMMARY ===\n")
ibd_spatial %>% count(deviation_type) %>% print()

# Aggregate barriers/corridors by region pair
barriers <- ibd_spatial %>%
  filter(str_detect(deviation_type, "barrier")) %>%
  group_by(region_pair) %>%
  summarise(n = n(), mean_fst = round(mean(Fst), 3),
            mean_resid = round(mean(residual), 3), .groups = "drop") %>%
  arrange(desc(mean_resid))

corridors <- ibd_spatial %>%
  filter(str_detect(deviation_type, "corridor")) %>%
  group_by(region_pair) %>%
  summarise(n = n(), mean_fst = round(mean(Fst), 3),
            mean_resid = round(mean(residual), 3), .groups = "drop") %>%
  arrange(mean_resid)

cat("\n=== BARRIERS (more isolated than expected from distance) ===\n")
print(barriers)
cat("\n=== CORRIDORS (more connected than expected from distance) ===\n")
print(corridors)

write_csv(barriers, "outputs/02_barriers.csv")
write_csv(corridors, "outputs/02_corridors.csv")
```

# Moderator Analysis and Variance Partitioning

## Marker type as moderator

```{r moderator-marker}
# Filter to comparisons with known marker type
ibd_markers <- fst_between %>% filter(!is.na(marker_category), marker_category != "Other")
cat("Comparisons with known marker:", nrow(ibd_markers), "\n")
ibd_markers %>% count(marker_category, sort = TRUE) %>% print()

# Model 1: Base (distance only)
mod_base <- lmer(Fst ~ log10(geo_dist_km) + (1 | study), data = ibd_markers)
r2_base <- r.squaredGLMM(mod_base)
vc_base <- as.data.frame(VarCorr(mod_base))
icc_base <- vc_base$vcov[vc_base$grp == "study"] / sum(vc_base$vcov)

cat("\n=== MODEL 1: Distance only ===\n")
cat("R²m:", round(r2_base[1, "R2m"], 4), " R²c:", round(r2_base[1, "R2c"], 4), "\n")
cat("ICC(study):", round(icc_base, 3), "\n")

# Model 2: + marker category
mod_marker <- lmer(Fst ~ log10(geo_dist_km) + marker_category + (1 | study), data = ibd_markers)
r2_marker <- r.squaredGLMM(mod_marker)
vc_marker <- as.data.frame(VarCorr(mod_marker))
icc_marker <- vc_marker$vcov[vc_marker$grp == "study"] / sum(vc_marker$vcov)

cat("\n=== MODEL 2: + marker type ===\n")
cat("R²m:", round(r2_marker[1, "R2m"], 4), " R²c:", round(r2_marker[1, "R2c"], 4), "\n")
cat("ICC(study):", round(icc_marker, 3), "\n")
cat("ICC reduction:", round(icc_base - icc_marker, 3),
    "(", round((icc_base - icc_marker) / icc_base * 100, 1), "% of study variance)\n")

cat("\nFixed effects:\n")
print(summary(mod_marker)$coefficients)

cat("\nLikelihood ratio test (Model 1 vs Model 2):\n")
print(anova(mod_base, mod_marker))

# Model 3: + species random effect
mod_species <- lmer(Fst ~ log10(geo_dist_km) + marker_category + (1 | study) + (1 | Species_A),
                    data = ibd_markers)
r2_species <- r.squaredGLMM(mod_species)
cat("\n=== MODEL 3: + species random effect ===\n")
cat("R²m:", round(r2_species[1, "R2m"], 4), " R²c:", round(r2_species[1, "R2c"], 4), "\n")
cat("Variance components:\n")
print(VarCorr(mod_species))
if (isSingular(mod_species)) cat("WARNING: Singular fit — Species_A explains ~0 variance\n")
```

## Descriptive statistics by marker type

```{r marker-descriptives}
marker_summary <- ibd_markers %>%
  group_by(marker_category) %>%
  summarise(
    n_comparisons = n(), n_studies = n_distinct(study),
    n_species = n_distinct(Species_A),
    mean_fst = round(mean(Fst), 4), sd_fst = round(sd(Fst), 4),
    median_fst = round(median(Fst), 4),
    mean_dist_km = round(mean(geo_dist_km)),
    .groups = "drop"
  ) %>%
  arrange(desc(n_comparisons))

print(marker_summary, width = Inf)

kw <- kruskal.test(Fst ~ marker_category, data = ibd_markers)
cat("\nKruskal-Wallis: H =", round(kw$statistic, 2), " p =", format(kw$p.value, digits = 3), "\n")

if (kw$p.value < 0.05) {
  cat("\nPairwise Wilcoxon (Bonferroni):\n")
  print(pairwise.wilcox.test(ibd_markers$Fst, ibd_markers$marker_category,
                             p.adjust.method = "bonferroni"))
}

write_csv(marker_summary, "outputs/03_marker_summary.csv")
```

```{r marker-boxplot, fig.width=6.7, fig.height=5}
# Order marker types by median Fst (ascending = highest connectivity first)
marker_order <- ibd_markers %>%
  group_by(marker_category) %>%
  summarise(med = median(Fst, na.rm = TRUE), .groups = "drop") %>%
  arrange(med) %>%
  pull(marker_category)

p_marker_box <- ibd_markers %>%
  mutate(marker_category = factor(marker_category, levels = marker_order)) %>%
  ggplot(aes(x = marker_category, y = Fst, fill = marker_category)) +
  geom_boxplot(outlier.shape = 21, outlier.size = 1.2, outlier.alpha = 0.4,
               width = 0.55, linewidth = 0.4) +
  geom_jitter(width = 0.14, size = 0.7, alpha = 0.25, colour = "grey20") +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2.5,
               fill = "white", colour = "black") +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(x = "Marker type", y = expression(F[ST]),
       title = expression(bold(F[ST])~"by marker type")) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor    = element_blank(),
    panel.grid.major.x  = element_blank(),
    axis.title          = element_text(size = 13, face = "bold"),
    axis.text           = element_text(size = 11),
    plot.title          = element_text(size = 12, face = "bold")
  )

print(p_marker_box)
ggsave("outputs/Figure_marker_boxplot.pdf", p_marker_box,
       width = 160, height = 130, units = "mm", dpi = 600, bg = "white")
```

## Variance partitioning

```{r variance-partitioning}
# Use ALL comparisons (between + within) with valid spawn mode
fst_all <- fst_full %>%
  filter(!is.na(Lat_A), !is.na(Lon_A), !is.na(Lat_B), !is.na(Lon_B)) %>%
  mutate(
    Lat_A = coalesce(as.numeric(as.character(Lat_A))),
    Lon_A = coalesce(as.numeric(as.character(Lon_A))),
    Lat_B = coalesce(as.numeric(as.character(Lat_B))),
    Lon_B = coalesce(as.numeric(as.character(Lon_B)))
  ) %>%
  rowwise() %>%
  mutate(geo_dist_km = distHaversine(c(Lon_A, Lat_A), c(Lon_B, Lat_B)) / 1000) %>%
  ungroup() %>%
  filter(geo_dist_km > 0) %>%
   mutate(
    spawn_mode = coalesce(Spawn_A, Spawn_B)
  ) %>%
  filter(!is.na(marker_category), marker_category != "Other",
         !is.na(spawn_mode), !is.na(Taxon_A))

cat("Dataset for variance partitioning:", nrow(fst_all), "comparisons\n")

# Progressive model building
vp_study     <- lmer(Fst ~ log10(geo_dist_km) + (1 | study), data = fst_all)
vp_taxon     <- lmer(Fst ~ log10(geo_dist_km) + (1 | study) + (1 | Taxon_A), data = fst_all)
vp_species   <- lmer(Fst ~ log10(geo_dist_km) + (1 | study) + (1 | Species_A), data = fst_all)
vp_spawn     <- lmer(Fst ~ log10(geo_dist_km) + (1 | study) + (1 | spawn_mode), data = fst_all)
vp_tax_mark  <- lmer(Fst ~ log10(geo_dist_km) + marker_category + (1 | study) + (1 | Taxon_A),
                     data = fst_all)
vp_full      <- lmer(Fst ~ log10(geo_dist_km) + marker_category +
                       (1 | study) + (1 | Taxon_A) + (1 | spawn_mode), data = fst_all)

# Extract variance proportions
get_vp <- function(model, name) {
  vc <- as.data.frame(VarCorr(model))
  total <- sum(vc$vcov)
  tibble(
    Model = name,
    Pct_Study   = round(100 * ifelse("study"      %in% vc$grp, vc$vcov[vc$grp == "study"], 0) / total, 1),
    Pct_Taxon   = round(100 * ifelse("Taxon_A"    %in% vc$grp, vc$vcov[vc$grp == "Taxon_A"], 0) / total, 1),
    Pct_Species = round(100 * ifelse("Species_A"  %in% vc$grp, vc$vcov[vc$grp == "Species_A"], 0) / total, 1),
    Pct_Spawn   = round(100 * ifelse("spawn_mode" %in% vc$grp, vc$vcov[vc$grp == "spawn_mode"], 0) / total, 1),
    Pct_Residual = round(100 * vc$vcov[vc$grp == "Residual"] / total, 1),
    Singular = isSingular(model)
  )
}

vp_table <- bind_rows(
  get_vp(vp_study,    "1. Study only"),
  get_vp(vp_taxon,    "2. + Taxon"),
  get_vp(vp_species,  "3. + Species"),
  get_vp(vp_spawn,    "4. + Spawn mode"),
  get_vp(vp_tax_mark, "5. + Taxon + Marker"),
  get_vp(vp_full,     "6. Full model")
)

cat("\n=== VARIANCE PARTITIONING (% of total) ===\n")
print(vp_table, width = Inf)

# Likelihood ratio tests
cat("\n=== LIKELIHOOD RATIO TESTS ===\n")
cat("Study vs Study+Taxon:\n");    print(anova(vp_study, vp_taxon))
cat("Study vs Study+Species:\n");  print(anova(vp_study, vp_species))
cat("Study vs Study+Spawn:\n");    print(anova(vp_study, vp_spawn))
cat("Study+Taxon vs +Marker:\n");  print(anova(vp_taxon, vp_tax_mark))
cat("+Marker vs Full:\n");         print(anova(vp_tax_mark, vp_full))

# Final variance table for manuscript
vc_final <- as.data.frame(VarCorr(vp_full))
total_var <- sum(vc_final$vcov)
variance_table_final <- tibble(
  Source = case_when(
    vc_final$grp == "study"      ~ "Study (methodological)",
    vc_final$grp == "Taxon_A"    ~ "Taxon (biological)",
    vc_final$grp == "spawn_mode" ~ "Spawn mode (reproductive)",
    vc_final$grp == "Residual"   ~ "Residual"
  ),
  Variance = round(vc_final$vcov, 6),
  Percent  = round(100 * vc_final$vcov / total_var, 1)
)
cat("\n=== FINAL VARIANCE TABLE ===\n")
print(variance_table_final)

write_csv(vp_table, "outputs/03_variance_partitioning.csv")
write_csv(variance_table_final, "outputs/03_variance_table_final.csv")
```

## Meta-analytic heterogeneity (I² and Cochran's Q)

```{r heterogeneity}
study_level <- fst_between %>%
  group_by(study) %>%
  summarise(
    n = n(), mean_fst = mean(Fst), sd_fst = sd(Fst),
    se_fst = sd_fst / sqrt(n),
    marker = first(marker_category), species = first(Species_A),
    .groups = "drop"
  ) %>%
  filter(!is.na(se_fst), se_fst > 0)

meta_re <- rma(yi = mean_fst, sei = se_fst, data = study_level, method = "REML")

cat("=== META-ANALYTIC HETEROGENEITY ===\n")
cat("Studies:", nrow(study_level), "\n")
cat("Cochran's Q:", round(meta_re$QE, 2), " p =", format(meta_re$QEp, digits = 4), "\n")
cat("I² =", round(meta_re$I2, 1), "%\n")
cat("H² =", round(meta_re$H2, 2), "\n")
cat("τ² =", round(meta_re$tau2, 5), "\n")

heterogeneity_stats <- tibble(
  Statistic = c("Cochran's Q", "Q p-value", "I²", "H²", "τ²"),
  Value = c(round(meta_re$QE, 2), format(meta_re$QEp, digits = 4, scientific = TRUE),
            paste0(round(meta_re$I2, 1), "%"), round(meta_re$H2, 2), round(meta_re$tau2, 5))
)
write_csv(heterogeneity_stats, "outputs/03_heterogeneity.csv")
```

## Sensitivity analyses

```{r sensitivity}
# Leave-one-out study influence
full_slope <- fixef(mod_base)[2]
full_r2m   <- r.squaredGLMM(mod_base)[1, "R2m"]

loo_results <- tibble(
  excluded = "NONE (full model)", n_remaining = nrow(ibd_markers),
  slope = round(full_slope, 4), r2m = round(full_r2m, 4),
  slope_pct_change = 0, r2_pct_change = 0, influential = FALSE
)

for (s in unique(ibd_markers$study)) {
  d <- ibd_markers %>% filter(study != s)
  if (nrow(d) < 50) next
  m <- tryCatch(lmer(Fst ~ log10(geo_dist_km) + (1 | study), data = d), error = function(e) NULL)
  if (is.null(m)) next
  sl <- fixef(m)[2]
  r2 <- r.squaredGLMM(m)[1, "R2m"]
  loo_results <- bind_rows(loo_results, tibble(
    excluded = s, n_remaining = nrow(d),
    slope = round(sl, 4), r2m = round(r2, 4),
    slope_pct_change = round((sl - full_slope) / full_slope * 100, 1),
    r2_pct_change = round((r2 - full_r2m) / full_r2m * 100, 1),
    influential = abs((sl - full_slope) / full_slope * 100) > 20 |
                  abs((r2 - full_r2m) / full_r2m * 100) > 20
  ))
}

cat("=== LEAVE-ONE-OUT SENSITIVITY ===\n")
print(loo_results, width = Inf)

# FST threshold sensitivity
threshold_sens <- expand_grid(low = c(0.02, 0.05, 0.07), high = c(0.10, 0.15, 0.20)) %>%
  rowwise() %>%
  mutate(data = list({
    region_connectivity %>%
      mutate(cat = case_when(mean_fst < low ~ "High", mean_fst < high ~ "Moderate", TRUE ~ "Low")) %>%
      count(cat) %>% pivot_wider(names_from = cat, values_from = n, values_fill = 0)
  })) %>%
  unnest(data) %>%
  mutate(total = High + Moderate + Low,
         pct_high = round(100 * High / total, 1),
         pct_mod  = round(100 * Moderate / total, 1),
         pct_low  = round(100 * Low / total, 1)) %>%
  ungroup()

cat("\n=== THRESHOLD SENSITIVITY ===\n")
print(threshold_sens)

write_csv(loo_results, "outputs/03_leave_one_out.csv")
write_csv(threshold_sens, "outputs/03_threshold_sensitivity.csv")
```

```{r threshold-sensitivity-plot, fig.width=8, fig.height=4.5}
thresh_long <- threshold_sens %>%
  select(low, high, pct_high, pct_mod, pct_low) %>%
  pivot_longer(c(pct_high, pct_mod, pct_low),
               names_to = "class", values_to = "pct") %>%
  mutate(
    class = factor(
      case_when(class == "pct_high" ~ "High",
                class == "pct_mod"  ~ "Moderate",
                TRUE                ~ "Low"),
      levels = c("Low", "Moderate", "High")
    ),
    facet_label = paste0("Low threshold: Fst < ", low),
    high_label  = paste0(high)
  )

# Mark the default thresholds (low = 0.05, high = 0.15)
default_bar <- thresh_long %>%
  filter(low == 0.05, high == 0.15) %>%
  select(facet_label, high_label) %>%
  distinct()

p_thresh <- ggplot(thresh_long,
                   aes(x = factor(high_label), y = pct, fill = class)) +
  geom_col(position = "stack", width = 0.65,
           colour = "white", linewidth = 0.3) +
  geom_rect(data = default_bar,
            aes(xmin = as.numeric(factor(high_label,
                  levels = unique(thresh_long$high_label))) - 0.38,
                xmax = as.numeric(factor(high_label,
                  levels = unique(thresh_long$high_label))) + 0.38,
                ymin = -2, ymax = 102),
            fill = NA, colour = "black", linewidth = 0.7,
            inherit.aes = FALSE) +
  facet_wrap(~facet_label, nrow = 1) +
  scale_fill_manual(
    values = c("High" = "firebrick", "Moderate" = "coral2", "Low" = "skyblue2"),
    name = "Connectivity"
  ) +
  scale_y_continuous(limits = c(0, 100), expand = c(0, 0)) +
  labs(x = expression("High connectivity threshold (F"[ST]*")"),
       y = "Region pairs (%)",
       title = "Sensitivity of connectivity classification to Fst thresholds",
       subtitle = "Black outline = default thresholds (Fst < 0.05 / > 0.15)") +
  theme_bw(base_size = 11) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title         = element_text(size = 12, face = "bold"),
    axis.text          = element_text(size = 10),
    strip.text         = element_text(size = 10, face = "bold"),
    legend.title       = element_text(size = 10, face = "bold"),
    legend.text        = element_text(size = 9),
    plot.title         = element_text(size = 11, face = "bold"),
    plot.subtitle      = element_text(size = 9, colour = "grey40")
  )

print(p_thresh)
ggsave("outputs/Figure_threshold_sensitivity.pdf", p_thresh,
       width = 200, height = 120, units = "mm", dpi = 600, bg = "white")
```

## Within-study Mantel tests

```{r mantel-by-study}
studies_for_mantel <- fst_between %>% count(study) %>% filter(n >= 10)

run_study_mantel <- function(study_id, data) {
  sd <- data %>% filter(study == study_id)
  sites <- bind_rows(
    sd %>% select(site = Site_A, lat = Lat_A, lon = Lon_A),
    sd %>% select(site = Site_B, lat = Lat_B, lon = Lon_B)
  ) %>% distinct(site, .keep_all = TRUE) %>% filter(!is.na(lat), !is.na(lon))

  ns <- nrow(sites)
  if (ns < 4) return(tibble(study = study_id, n_sites = ns, mantel_r = NA, p = NA, note = "<4 sites"))

  crds <- as.matrix(sites[, c("lon", "lat")]); rownames(crds) <- sites$site
  gm <- tryCatch(distm(crds, fun = distHaversine) / 1000, error = function(e) NULL)
  if (is.null(gm)) return(tibble(study = study_id, n_sites = ns, mantel_r = NA, p = NA, note = "geo error"))
  rownames(gm) <- colnames(gm) <- sites$site

  fm <- matrix(NA, ns, ns, dimnames = list(sites$site, sites$site))
  for (i in seq_len(nrow(sd))) {
    a <- sd$Site_A[i]; b <- sd$Site_B[i]
    if (a %in% rownames(fm) & b %in% colnames(fm)) { fm[a, b] <- sd$Fst[i]; fm[b, a] <- sd$Fst[i] }
  }
  diag(fm) <- 0

  comp <- sum(!is.na(fm[upper.tri(fm)])) / sum(upper.tri(fm))
  if (comp < 0.5) return(tibble(study = study_id, n_sites = ns, mantel_r = NA, p = NA,
                                note = paste0(round(comp * 100), "% complete")))

  n_imp <- sum(is.na(fm[upper.tri(fm)]))
  if (any(is.na(fm))) fm[is.na(fm)] <- mean(fm, na.rm = TRUE)

  set.seed(42)
  mt <- tryCatch(mantel(as.dist(fm), as.dist(gm), method = "spearman", permutations = 999),
                 error = function(e) NULL)
  if (is.null(mt)) return(tibble(study = study_id, n_sites = ns, mantel_r = NA, p = NA, note = "failed"))

  tibble(study = study_id, n_sites = ns, mantel_r = round(mt$statistic, 4),
         p = round(mt$signif, 4), note = ifelse(n_imp > 0, paste(n_imp, "imputed"), "complete"))
}

mantel_by_study <- map_dfr(studies_for_mantel$study, run_study_mantel, data = fst_between) %>%
  left_join(fst_between %>% group_by(study) %>%
              summarise(marker = first(marker_category), species = first(Species_A),
                        n_pairs = n(), .groups = "drop"),
            by = "study")

cat("=== WITHIN-STUDY MANTEL TESTS ===\n")
print(mantel_by_study, width = Inf)
write_csv(mantel_by_study, "outputs/03_mantel_by_study.csv")
```

# Genetic Network Analysis

## Compute genetic centrality

```{r genetic-network}
# Force edge attribute columns to plain numeric vectors to avoid list-column errors
fst_vals  <- as.numeric(region_connectivity$mean_fst)
ncomp_vals <- as.integer(region_connectivity$n_comparisons)
conn_vals  <- as.character(region_connectivity$connectivity)
weight_vals <- pmax(1 - fst_vals, 1e-6, na.rm = TRUE)  # guard against 0/negative weights

# Build undirected weighted graph from region-pair data
g_gen <- graph_from_data_frame(
  region_connectivity %>% select(Region_A, Region_B),
  directed = FALSE
)
E(g_gen)$weight       <- weight_vals
E(g_gen)$fst          <- fst_vals
E(g_gen)$n_comp       <- ncomp_vals
E(g_gen)$connectivity <- conn_vals

genetic_centrality <- tibble(
  region              = V(g_gen)$name,
  genetic_degree      = as.numeric(igraph::degree(g_gen)),
  genetic_strength    = round(as.numeric(strength(g_gen, weights = E(g_gen)$weight)), 2),
  genetic_betweenness = round(as.numeric(betweenness(g_gen, weights = 1 / E(g_gen)$weight)), 1)
) %>%
  arrange(desc(genetic_degree))

cat("=== GENETIC NETWORK CENTRALITY ===\n")
print(genetic_centrality, width = Inf)
write_csv(genetic_centrality, "outputs/04_genetic_centrality.csv")
```

```{r genetic-network-map, fig.width=7.1, fig.height=6.3}
# region_centroids loaded from regionData.csv in load-data chunk

net_edges <- region_connectivity %>%
  left_join(region_centroids, by = c("Region_A" = "region")) %>%
  rename(lon_A = lon, lat_A = lat) %>%
  left_join(region_centroids, by = c("Region_B" = "region")) %>%
  rename(lon_B = lon, lat_B = lat) %>%
  filter(!is.na(lon_A), !is.na(lat_A), !is.na(lon_B), !is.na(lat_B)) %>%
  arrange(connectivity)

conn_cols <- c("Low" = "skyblue2", "Moderate" = "coral2", "High" = "firebrick")

# Shared publication theme — applied to both network maps
map_theme <- theme_bw(base_size = 12) +
  theme(
    panel.grid       = element_blank(),
    panel.background = element_rect(fill = "aliceblue"),
    axis.title       = element_text(size = 13, face = "bold"),
    axis.text        = element_text(size = 11, colour = "black"),
    legend.title     = element_text(size = 9, face = "bold"),
    legend.text      = element_text(size = 8),
    legend.key.size  = unit(0.4, "cm"),
    legend.position  = "right",
    plot.title       = element_text(size = 14, face = "bold", hjust = 0)
  )

# Shared inset theme
inset_theme_base <- theme_bw(base_size = 9) +
  theme(
    panel.grid       = element_blank(),
    panel.background = element_rect(fill = "aliceblue"),
    panel.border     = element_rect(colour = "black", linewidth = 0.8),
    axis.title       = element_blank(),
    axis.text        = element_blank(),
    axis.ticks       = element_blank(),
    legend.position  = "none",
    plot.margin      = margin(0, 0, 0, 0),
    plot.background  = element_rect(fill = "white", colour = "black", linewidth = 0.6)
  )

# Inset bounding box
inset_xlim <- c(113, 117.5)
inset_ylim <- c(-23.5, -20)

# Convert to sf for accurate node placement with coord_sf
region_centroids_sf <- st_as_sf(region_centroids, coords = c("lon", "lat"),
                                crs = 4326, remove = FALSE)
inset_centroids_sf <- region_centroids_sf %>%
  filter(lon >= inset_xlim[1], lon <= inset_xlim[2],
         lat >= inset_ylim[1], lat <= inset_ylim[2])

# ── Main map (with dashed locator box) ────────────────────────────────────────
p_gen_net <- ggplot() +
  geom_sf(data = basemap, fill = "grey80", color = "black", linewidth = 0.25) +
  geom_segment(data = net_edges,
    aes(x = lon_A, y = lat_A, xend = lon_B, yend = lat_B, colour = connectivity),
    linewidth = 1.0, alpha = 0.75) +
  geom_sf(data = region_centroids_sf, size = 3, shape = 19,
          colour = "black", inherit.aes = FALSE) +
  annotate("rect",
    xmin = inset_xlim[1], xmax = inset_xlim[2],
    ymin = inset_ylim[1], ymax = inset_ylim[2],
    colour = "black", fill = NA, linewidth = 0.6, linetype = "dashed") +
  scale_colour_manual(values = conn_cols,
                      name = "Genetic\nConnectivity") +
  annotation_scale(location = "bl", width_hint = 0.15, style = "ticks",
                   text_cex = 0.8, line_width = 0.6) +
  coord_sf(xlim = c(108, 130), ylim = c(-36, -8), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude", title = "(a)") +
  map_theme

# ── Inset ─────────────────────────────────────────────────────────────────────
gen_inset_base <- list(
  geom_sf(data = basemap, fill = "grey85", color = "grey50", linewidth = 0.25),
  geom_segment(data = net_edges,
    aes(x = lon_A, y = lat_A, xend = lon_B, yend = lat_B, colour = connectivity),
    linewidth = 0.9, alpha = 0.75),
  geom_sf(data = region_centroids_sf, size = 2, shape = 19,
          colour = "black", inherit.aes = FALSE),
  scale_colour_manual(values = conn_cols, guide = "none"),
  coord_sf(xlim = inset_xlim, ylim = inset_ylim, expand = FALSE),
  inset_theme_base
)

p_gen_inset <- ggplot() + gen_inset_base

# ── Save main map and inset as separate files ─────────────────────────────────
print(p_gen_net)
ggsave("outputs/Figure_genetic_network.pdf", p_gen_net,
       width = 180, height = 160, units = "mm", dpi = 600, bg = "white")
ggsave("outputs/Figure_genetic_network.png", p_gen_net,
       width = 180, height = 160, units = "mm", dpi = 300, bg = "white")

print(p_gen_inset)
ggsave("outputs/Figure_genetic_network_inset.pdf", p_gen_inset,
       width = 80, height = 80, units = "mm", dpi = 600, bg = "white")
ggsave("outputs/Figure_genetic_network_inset.png", p_gen_inset,
       width = 80, height = 80, units = "mm", dpi = 300, bg = "white")
```

# Hydrodynamic Network Analysis

## Compute hydrodynamic centrality

```{r hydro-network}
g_hydro <- graph_from_data_frame(
  hydro_summary %>%
    mutate(from = sapply(str_split(region_pair, " <-> "), `[`, 1),
           to   = sapply(str_split(region_pair, " <-> "), `[`, 2)) %>%
    select(from, to),
  directed = FALSE
)
E(g_hydro)$weight <- hydro_summary$hydro_strength_num
E(g_hydro)$consensus <- hydro_summary$hydro_consensus

hydro_centrality <- tibble(
  region              = V(g_hydro)$name,
  hydro_degree        = as.numeric(igraph::degree(g_hydro)),
  hydro_strength      = round(as.numeric(strength(g_hydro, weights = E(g_hydro)$weight)), 2),
  hydro_betweenness   = round(as.numeric(betweenness(g_hydro, weights = 1 / E(g_hydro)$weight)), 1)
) %>%
  arrange(desc(hydro_degree))

cat("=== HYDRODYNAMIC NETWORK CENTRALITY ===\n")
print(hydro_centrality, width = Inf)
write_csv(hydro_centrality, "outputs/05_hydro_centrality.csv")

# Directional summary for manuscript
cat("\n=== DIRECTIONAL TRANSPORT SUMMARY ===\n")
cat("Total connections:", nrow(hydro_conn), "\n")
hydro_conn %>%
  mutate(dir_class = case_when(
    direction %in% c("S", "SW", "SE") ~ "Southward/SW/SE",
    direction == "W" ~ "Westward",
    direction == "bidirectional" ~ "Bidirectional",
    TRUE ~ "Other"
  )) %>%
  count(dir_class) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  print()
```

```{r hydro-network-map, fig.width=7.1, fig.height=6.3}
# Parse region pairs into from/to and join coordinates
hydro_edges <- hydro_summary %>%
  mutate(
    from = sapply(str_split(region_pair, " <-> "), `[`, 1),
    to   = sapply(str_split(region_pair, " <-> "), `[`, 2),
    is_bidirectional = str_detect(directions, "bidirectional"),
    hydro_consensus  = factor(hydro_consensus, levels = c("low", "mod", "high"))
  ) %>%
  left_join(region_centroids, by = c("from" = "region")) %>%
  rename(lon_A = lon, lat_A = lat) %>%
  left_join(region_centroids, by = c("to" = "region")) %>%
  rename(lon_B = lon, lat_B = lat) %>%
  filter(!is.na(lon_A), !is.na(lat_A), !is.na(lon_B), !is.na(lat_B)) %>%
  arrange(hydro_consensus)

hydro_cols <- c("low" = "skyblue2", "mod" = "coral2", "high" = "firebrick")
hydro_labs <- c("low" = "Low", "mod" = "Moderate", "high" = "High")

hydro_bidir <- filter(hydro_edges, is_bidirectional)

# Directed: orient north→south, then shorten destination end to clear node
hydro_directed <- filter(hydro_edges, !is_bidirectional) %>%
  mutate(
    # Swap so x_start/y_start is always the northern (higher lat) node
    going_south = lat_A >= lat_B,
    x_start = ifelse(going_south, lon_A, lon_B),
    y_start = ifelse(going_south, lat_A, lat_B),
    x_end   = ifelse(going_south, lon_B, lon_A),
    y_end   = ifelse(going_south, lat_B, lat_A)
  ) %>%
  mutate(
    dx    = x_end - x_start,
    dy    = y_end - y_start,
    d     = sqrt(dx^2 + dy^2),
    x_end = x_end - 0.35 * dx / d,
    y_end = y_end - 0.35 * dy / d
  )

# ── Main map ───────────────────────────────────────────────────────────────────
p_hydro_net <- ggplot() +
  geom_sf(data = basemap, fill = "grey85", color = "grey50", linewidth = 0.25) +
  geom_segment(data = hydro_bidir,
    aes(x = lon_A, y = lat_A, xend = lon_B, yend = lat_B, colour = hydro_consensus),
    linewidth = 1.0, alpha = 0.75) +
  geom_segment(data = hydro_directed,
    aes(x = x_start, y = y_start, xend = x_end, yend = y_end, colour = hydro_consensus),
    linewidth = 1.0, alpha = 0.75,
    arrow = arrow(length = unit(0.2, "cm"), type = "closed")) +
  geom_sf(data = region_centroids_sf, size = 3, shape = 19,
          colour = "black", inherit.aes = FALSE) +
  scale_colour_manual(values = hydro_cols, labels = hydro_labs,
                      breaks = c("low", "mod", "high"),
                      name = "Hydrodynamic\nConnectivity") +
  annotation_scale(location = "bl", width_hint = 0.15, style = "ticks",
                   text_cex = 0.8, line_width = 0.6) +
  coord_sf(xlim = c(108, 130), ylim = c(-36, -8), expand = FALSE) +
  labs(x = "Longitude", y = "Latitude", title = "(b)") +
  map_theme

print(p_hydro_net)
ggsave("outputs/Figure_hydro_network.pdf", p_hydro_net,
       width = 180, height = 160, units = "mm", dpi = 600, bg = "white")
```

## Connectivity heatmaps

```{r connectivity-heatmaps, fig.width=14, fig.height=8}
# Region order: north→south (from regionData.csv)
region_order <- region_centroids$region

# Helper: build symmetric region × region matrix
make_sym_mat <- function(df, col_A, col_B, col_val, regions) {
  n   <- length(regions)
  mat <- matrix(NA_real_, n, n, dimnames = list(regions, regions))
  for (i in seq_len(nrow(df))) {
    a <- df[[col_A]][i]; b <- df[[col_B]][i]; v <- df[[col_val]][i]
    if (!is.na(a) && !is.na(b) && a %in% regions && b %in% regions) {
      mat[a, b] <- v
      mat[b, a] <- v
    }
  }
  mat
}

# Genetic Fst matrix
fst_mat <- make_sym_mat(
  region_connectivity, "Region_A", "Region_B", "mean_fst", region_order
)

# Hydrodynamic strength matrix
hydro_df <- hydro_summary %>%
  mutate(
    R_A = sapply(str_split(region_pair, " <-> "), `[`, 1),
    R_B = sapply(str_split(region_pair, " <-> "), `[`, 2)
  )
hydro_mat <- make_sym_mat(
  hydro_df, "R_A", "R_B", "hydro_strength_num", region_order
)

# Colour scales
fst_max  <- ceiling(max(fst_mat, na.rm = TRUE) * 20) / 20
fst_col  <- colorRampPalette(c("#2166ac", "#f7f7f7", "#d73027"))(100)
fst_brks <- seq(0, fst_max, length.out = 101)

hydro_col  <- c("skyblue2", "coral2", "firebrick")
hydro_brks <- c(0.5, 1.5, 2.5, 3.5)

# Genetic heatmap
ph_fst <- pheatmap(
  fst_mat,
  color         = fst_col,
  breaks        = fst_brks,
  na_col        = "grey92",
  cluster_rows  = FALSE, cluster_cols = FALSE,
  show_rownames = TRUE,  show_colnames = TRUE,
  angle_col     = 45,
  fontsize      = 8,
  main          = "(a) Genetic Connectivity (mean Fst)",
  silent        = TRUE
)

# Hydrodynamic heatmap
ph_hydro <- pheatmap(
  hydro_mat,
  color         = hydro_col,
  breaks        = hydro_brks,
  legend_breaks = c(1, 2, 3),
  legend_labels = c("Low", "Moderate", "High"),
  na_col        = "grey92",
  cluster_rows  = FALSE, cluster_cols = FALSE,
  show_rownames = TRUE,  show_colnames = TRUE,
  angle_col     = 45,
  fontsize      = 8,
  main          = "(b) Hydrodynamic Connectivity",
  silent        = TRUE
)

# Combine side-by-side and save
p_heatmaps <- cowplot::plot_grid(ph_fst$gtable, ph_hydro$gtable, nrow = 1)
print(p_heatmaps)

ggsave("outputs/Figure_heatmaps.pdf", p_heatmaps,
       width = 360, height = 200, units = "mm", dpi = 600, bg = "white")
```

# Genetic–Hydrodynamic Concordance

## Region-pair concordance

```{r concordance}
# Merge genetic and hydrodynamic at region-pair level
gen_for_merge <- region_connectivity %>%
  select(region_pair, mean_fst, n_comparisons, connectivity) %>%
  rename(genetic_connectivity = connectivity) %>%
  mutate(genetic_strength_num = case_when(
    genetic_connectivity == "High" ~ 3L,
    genetic_connectivity == "Moderate" ~ 2L,
    genetic_connectivity == "Low" ~ 1L
  ))

concordance <- gen_for_merge %>%
  inner_join(hydro_summary, by = "region_pair")

cat("=== REGION PAIRS WITH BOTH DATASETS ===\n")
cat("Overlapping pairs:", nrow(concordance), "\n\n")
print(concordance %>%
        select(region_pair, mean_fst, genetic_connectivity, hydro_consensus,
               genetic_strength_num, hydro_strength_num),
      width = Inf)

# Jaccard similarity
gen_pairs   <- unique(gen_for_merge$region_pair)
hydro_pairs <- unique(hydro_summary$region_pair)
jaccard <- length(intersect(gen_pairs, hydro_pairs)) / length(union(gen_pairs, hydro_pairs))
cat("\n=== JACCARD SIMILARITY ===\n")
cat("Genetic pairs:", length(gen_pairs), "\n")
cat("Hydro pairs:", length(hydro_pairs), "\n")
cat("Overlap:", length(intersect(gen_pairs, hydro_pairs)), "\n")
cat("Union:", length(union(gen_pairs, hydro_pairs)), "\n")
cat("Jaccard =", round(jaccard, 3), "(", round(jaccard * 100, 1), "%)\n")

# Ordinal strength correlation (at region-pair level)
if (nrow(concordance) >= 5) {
  sp_ordinal <- cor.test(concordance$genetic_strength_num, concordance$hydro_strength_num,
                         method = "spearman")
  cat("\nSpearman (ordinal strength categories): rho =", round(sp_ordinal$estimate, 3),
      " p =", format(sp_ordinal$p.value, digits = 3), "\n")

  sp_fst <- cor.test(concordance$mean_fst, concordance$hydro_strength_num, method = "spearman")
  cat("Spearman (mean FST vs hydro strength): rho =", round(sp_fst$estimate, 3),
      " p =", format(sp_fst$p.value, digits = 3), "\n")

  cat("\nContingency table:\n")
  print(table(Genetic = concordance$genetic_connectivity, Hydro = concordance$hydro_consensus))
}
```

## Centrality comparison across networks

```{r centrality-comparison}
# Join centrality metrics for regions present in BOTH networks
centrality_comp <- genetic_centrality %>%
  inner_join(hydro_centrality, by = "region")

cat("=== CENTRALITY COMPARISON (regions in both networks) ===\n")
cat("Regions compared:", nrow(centrality_comp), "\n\n")
print(centrality_comp, width = Inf)

# Spearman correlations for each centrality metric
cat("\n=== SPEARMAN RANK CORRELATIONS ===\n")
for (metric in c("degree", "strength", "betweenness")) {
  g_col <- paste0("genetic_", metric)
  h_col <- paste0("hydro_", metric)
  sp <- cor.test(centrality_comp[[g_col]], centrality_comp[[h_col]], method = "spearman")
  cat(sprintf("%-12s: rho = %+.3f, p = %.3f (n = %d)\n",
              metric, sp$estimate, sp$p.value, nrow(centrality_comp)))
}

write_csv(centrality_comp, "outputs/06_centrality_comparison.csv")
write_csv(concordance, "outputs/06_concordance.csv")
```

```{r centrality-table}
# Full centrality table: all regions from both networks
all_centrality <- genetic_centrality %>%
  rename(Gen_Degree = genetic_degree,
         Gen_Strength = genetic_strength,
         Gen_Betweenness = genetic_betweenness) %>%
  full_join(
    hydro_centrality %>%
      rename(Hydro_Degree = hydro_degree,
             Hydro_Strength = hydro_strength,
             Hydro_Betweenness = hydro_betweenness),
    by = "region"
  ) %>%
  arrange(desc(coalesce(Gen_Degree, 0L) + coalesce(Hydro_Degree, 0L))) %>%
  rename(Region = region)

knitr::kable(
  all_centrality,
  digits  = c(0, 0, 2, 1, 0, 2, 1),
  col.names = c("Region",
                "Degree", "Strength", "Betweenness",
                "Degree", "Strength", "Betweenness"),
  caption = "Centrality metrics for genetic and hydrodynamic connectivity networks.
             Degree = number of connections; Strength = sum of weighted connections;
             Betweenness = number of shortest paths through node.
             NA = region absent from that network.",
  format  = "html"
) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                            full_width = FALSE, font_size = 12) %>%
  kableExtra::add_header_above(
    c(" " = 1, "Genetic network" = 3, "Hydrodynamic network" = 3)
  ) %>%
  kableExtra::column_spec(1, bold = TRUE)
```

## Concordance classification

```{r concordance-classification}
# Classify each overlapping pair
concordance_classified <- concordance %>%
  mutate(
    agreement = case_when(
      genetic_strength_num == hydro_strength_num           ~ "Concordant",
      abs(genetic_strength_num - hydro_strength_num) == 1  ~ "Partially concordant",
      TRUE                                                  ~ "Discordant"
    )
  )

cat("=== CONCORDANCE CLASSIFICATION ===\n")
concordance_classified %>% count(agreement) %>% print()
cat("\n")
concordance_classified %>%
  select(region_pair, genetic_connectivity, hydro_consensus, agreement) %>%
  print(width = Inf)

write_csv(concordance_classified, "outputs/06_concordance_classified.csv")
```

# Summary Tables for Manuscript

```{r summary-tables}
cat("================================================================\n")
cat("     MANUSCRIPT-READY SUMMARY — ALL KEY NUMBERS\n")
cat("================================================================\n\n")

cat("── DATA OVERVIEW ──\n")
cat("Total FST comparisons (cleaned):", nrow(fst_clean), "\n")
cat("Between-region:", nrow(fst_between), "\n")
cat("Within-region:", nrow(fst_within), "\n")
cat("Unique region pairs:", nrow(region_connectivity), "\n")
cat("Studies contributing FST data:", n_distinct(fst_between$study), "\n")
cat("Species:", n_distinct(fst_between$Species_A), "\n")
cat("Distance range:", round(min(fst_between$geo_dist_km)), "–",
    round(max(fst_between$geo_dist_km)), "km\n")

cat("\n── CONNECTIVITY CLASSIFICATION ──\n")
region_connectivity %>%
  count(connectivity) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  print()

cat("\n── ISOLATION-BY-DISTANCE ──\n")
cat("Simple LM: slope =", round(coef(ibd_lm)[2], 4), ", R² =", round(lm_summ$r.squared, 4), "\n")
cat("Bootstrap slope [95% BCa CI]:", round(boot_slope$t0, 4),
    "[", round(ci_slope$bca[4], 4), "–", round(ci_slope$bca[5], 4), "]\n")
cat("Mixed model: R²m =", round(r2_mixed[1, "R2m"], 4),
    ", R²c =", round(r2_mixed[1, "R2c"], 4), "\n")
cat("ICC(study) =", round(icc, 3), "\n")
cat("Moran's I =", round(moran5$estimate[1], 4), ", p =", format(moran5$p.value, digits = 3), "\n")
cat("Mantel r (log distance) =", round(mantel_log$statistic, 4), ", p =", mantel_log$signif, "\n")

cat("\n── MODERATOR ANALYSIS ──\n")
cat("Marker type LRT: χ² = 15.66, df = 3, p = 0.001\n")
cat("R²m increased from", round(r2_base[1, "R2m"], 4), "to", round(r2_marker[1, "R2m"], 4), "\n")
cat("ICC reduced from", round(icc_base, 3), "to", round(icc_marker, 3),
    "(", round((icc_base - icc_marker) / icc_base * 100, 1), "% of study variance explained)\n")
cat("Kruskal-Wallis: p =", format(kw$p.value, digits = 3), "\n")

cat("\n── HETEROGENEITY ──\n")
cat("Cochran's Q =", round(meta_re$QE, 2), ", p =", format(meta_re$QEp, digits = 3), "\n")
cat("I² =", round(meta_re$I2, 1), "%\n")

cat("\n── VARIANCE PARTITIONING (Full model) ──\n")
print(variance_table_final)

cat("\n── GENETIC NETWORK ──\n")
cat("Top hubs (degree):", paste(head(genetic_centrality$region, 3), collapse = ", "), "\n")
print(head(genetic_centrality, 5))

cat("\n── HYDRODYNAMIC NETWORK ──\n")
cat("Connections:", nrow(hydro_conn), "\n")
cat("Top hubs (degree):", paste(head(hydro_centrality$region, 3), collapse = ", "), "\n")
print(head(hydro_centrality, 5))

cat("\n── CONCORDANCE ──\n")
cat("Jaccard similarity:", round(jaccard * 100, 1), "%\n")
cat("Overlapping region pairs:", nrow(concordance), "\n")
cat("Spearman centrality correlations:\n")
for (metric in c("degree", "strength", "betweenness")) {
  sp <- cor.test(centrality_comp[[paste0("genetic_", metric)]],
                 centrality_comp[[paste0("hydro_", metric)]], method = "spearman")
  cat(sprintf("  %-12s: rho = %+.3f, p = %.3f\n", metric, sp$estimate, sp$p.value))
}

cat("\n── REGIONS WITHOUT DATA ──\n")
all_20_regions <- c("Timor Sea", "Ashmore", "Scott Reef", "Kimberley Nearshore",
                    "Kimberley Offshore", "Rowley Shoals", "NW nearshore", "NW offshore",
                    "Dampier Peninsula", "Dampier Archipelago", "Pilbara Offshore",
                    "Pilbara nearshore", "Exmouth Gulf", "Ningaloo", "Shark Bay",
                    "Abrolhos", "Central Coast", "Southwest coast",
                    "Cocos (Keeling)", "Christmas Island")
gen_regions <- unique(c(fst_between$Region_A, fst_between$Region_B))
hydro_regions <- unique(c(hydro_conn$region_A, hydro_conn$region_B))
either_regions <- union(gen_regions, hydro_regions)
uncharacterised <- setdiff(all_20_regions, either_regions)
cat("Characterised by genetics:", length(gen_regions), "\n")
cat("Characterised by hydro:", length(hydro_regions), "\n")
cat("Either approach:", length(either_regions), "\n")
cat("Uncharacterised (", length(uncharacterised), "):", paste(uncharacterised, collapse = ", "), "\n")
```

# Manuscript Tables

## Table S1 — Genetic studies reviewed

```{r ms-table-s1}
ms_gen <- litrev_genetic %>%
  arrange(publishedYear) %>%
  transmute(
    `Study ID`     = studyID,
    `Year`         = publishedYear,
    `Taxa`         = taxa,
    `Species`      = species,
    `Spawn mode`   = str_to_sentence(spawnType),
    `Marker`       = markerType,
    `Sites`        = numSites,
    `n (samples)`  = sampleSize,
    `Regions`      = location,
    `Included`     = if_else(valuesAvail == "y", "Yes", "No")
  )

kbl(ms_gen,
    caption = "Table S1. Genetic connectivity studies included in the literature review. Studies in grey were excluded from quantitative analysis (Included = No). n = total number of individual samples.",
    align   = c("l","c","l","l","l","l","c","c","l","c")) %>%
  kable_classic(full_width = FALSE, html_font = "Arial", font_size = 10) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(9, width = "18em") %>%
  row_spec(which(ms_gen$Included == "No"), color = "grey60") %>%
  row_spec(0, bold = TRUE)
```

## Table S2 — Dispersal model studies reviewed

```{r ms-table-s2}
ms_disp <- litrev_dispersal %>%
  arrange(publishedYear) %>%
  transmute(
    `Study ID`        = studyID,
    `Year`            = publishedYear,
    `Taxa`            = taxa,
    `Model`           = modelType,
    `Resolution`      = modelRes,
    `Season modelled` = seasonModel,
    `Tracking`        = trackingApproach,
    `Regions`         = location,
    `Included`        = if_else(valuesAvail == "y", "Yes", "No")
  )

kbl(ms_disp,
    caption = "Table S2. Hydrodynamic dispersal modelling studies included in the literature review. Studies in grey were excluded from quantitative analysis (Included = No).",
    align   = c("l","c","l","l","l","l","l","l","c")) %>%
  kable_classic(full_width = FALSE, html_font = "Arial", font_size = 10) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(8, width = "18em") %>%
  row_spec(which(ms_disp$Included == "No"), color = "grey60") %>%
  row_spec(0, bold = TRUE)
```

## Table 1 — Isolation-by-distance and heterogeneity

```{r ms-table-1}
ms_ibd <- tribble(
  ~Statistic,                            ~Value,                                                                                ~Note,
  # IBD - simple LM
  "IBD slope (simple LM)",               as.character(round(coef(ibd_lm)[2], 4)),                                               "FST per log10 km",
  "R2 (simple LM)",                      as.character(round(lm_summ$r.squared, 3)),                                             "",
  "Bootstrap slope [95% BCa CI]",        paste0(round(boot_slope$t0, 4), " [", round(ci_slope$bca[4], 4), " - ", round(ci_slope$bca[5], 4), "]"), "1,000 resamples",
  "Bootstrap R2 [95% BCa CI]",           paste0(round(boot_r2$t0, 3),    " [", round(ci_r2$bca[4],    3), " - ", round(ci_r2$bca[5],    3), "]"), "1,000 resamples",
  # IBD - mixed model
  "R2m marginal (distance)",             as.character(round(r2_mixed[1, "R2m"], 3)),                                            "Mixed model, study random intercept",
  "R2c conditional (distance + study)",  as.character(round(r2_mixed[1, "R2c"], 3)),                                            "",
  "ICC (study)",                         as.character(round(icc, 3)),                                                           "Proportion of variance among studies",
  # Spatial autocorrelation
  "Moran's I",                           as.character(round(moran5$estimate[1], 4)),                                            "k = 5 nearest neighbours",
  "Moran's I p-value",                   format(moran5$p.value, digits = 3, scientific = FALSE),                                "",
  "Mantel r (log distance)",             as.character(round(mantel_log$statistic, 4)),                                          "9,999 permutations, Spearman",
  "Mantel p-value",                      as.character(mantel_log$signif),                                                       "",
  # Heterogeneity
  "Cochran's Q",                         as.character(round(meta_re$QE, 2)),                                                    paste0("p = ", format(meta_re$QEp, digits = 3, scientific = TRUE)),
  "I2 (%)",                              paste0(round(meta_re$I2, 1), "%"),                                                     "Among-study heterogeneity",
  "tau2",                                as.character(round(meta_re$tau2, 5)),                                                   "Between-study variance"
)

kbl(ms_ibd,
    col.names = c("Statistic", "Value", "Notes"),
    caption   = "Table 1. Isolation-by-distance (IBD) and meta-analytic heterogeneity statistics for pairwise FST across Australian coral reef regions.",
    align     = c("l","c","l")) %>%
  kable_classic(full_width = FALSE, html_font = "Arial", font_size = 10) %>%
  row_spec(0, bold = TRUE) %>%
  pack_rows("Isolation-by-distance", 1, 7) %>%
  pack_rows("Spatial autocorrelation", 8, 11) %>%
  pack_rows("Heterogeneity", 12, 14)
```

## Table 2 — FST by marker type

```{r ms-table-2}
ms_marker <- marker_summary %>%
  transmute(
    `Marker type`    = marker_category,
    `Comparisons`    = n_comparisons,
    `Studies`        = n_studies,
    `Species`        = n_species,
    `Mean FST`       = mean_fst,
    `SD`             = sd_fst,
    `Median FST`     = median_fst,
    `Mean dist. (km)` = mean_dist_km
  )

# Kruskal-Wallis and pairwise results as a footnote string
kw_note <- paste0("Kruskal-Wallis H = ", round(kw$statistic, 2),
                  ", p = ", format(kw$p.value, digits = 3, scientific = FALSE))

kbl(ms_marker,
    caption = paste0("Table 2. FST summary statistics by molecular marker type. ", kw_note, "."),
    align   = c("l","c","c","c","c","c","c","c")) %>%
  kable_classic(full_width = FALSE, html_font = "Arial", font_size = 10) %>%
  row_spec(0, bold = TRUE) %>%
  footnote(general = kw_note, general_title = "")
```

## Table 3 — Variance partitioning

```{r ms-table-3}
r2m_vals <- c(
  r.squaredGLMM(vp_study)[1,    "R2m"],
  r.squaredGLMM(vp_taxon)[1,    "R2m"],
  r.squaredGLMM(vp_species)[1,  "R2m"],
  r.squaredGLMM(vp_spawn)[1,    "R2m"],
  r.squaredGLMM(vp_tax_mark)[1, "R2m"],
  r.squaredGLMM(vp_full)[1,     "R2m"]
)

ms_vp <- vp_table %>%
  mutate(R2m = round(r2m_vals, 3)) %>%
  transmute(
    Model          = Model,
    R2m            = R2m,
    `% Study`      = Pct_Study,
    `% Taxon`      = Pct_Taxon,
    `% Spawn mode` = Pct_Spawn,
    `% Residual`   = Pct_Residual,
    Singular       = if_else(Singular, "Yes", "")
  )

kbl(ms_vp,
    col.names = c("Model", "R2m (marginal)", "% Study", "% Taxon", "% Spawn mode", "% Residual", "Singular"),
    caption   = "Table 3. Progressive variance partitioning of FST. Values show the percentage of total variance attributed to each random effect. R2m = marginal R2 (fixed effects only). Marker type was included as a fixed effect in models 5-6.",
    align     = c("l","c","c","c","c","c","c")) %>%
  kable_classic(full_width = FALSE, html_font = "Arial", font_size = 10) %>%
  row_spec(0, bold = TRUE) %>%
  row_spec(nrow(ms_vp), bold = TRUE) %>%
  footnote(general = "Singular = model hit a boundary (variance component ≈ 0).",
           general_title = "")
```

## Table 4 — Network centrality and concordance

```{r ms-table-4}
# Centrality comparison table (regions in both networks)
ms_centrality <- all_centrality %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

kbl(ms_centrality,
    col.names = c("Region",
                  "Degree", "Strength", "Betweenness",
                  "Degree", "Strength", "Betweenness"),
    caption   = "Table 4. Network centrality metrics for genetic and hydrodynamic connectivity networks across Australian coral reef regions. Degree = number of connected region pairs; Strength = sum of edge weights; Betweenness = number of shortest paths through each node. NA = region absent from that network.",
    align     = c("l","c","c","c","c","c","c")) %>%
  kable_classic(full_width = FALSE, html_font = "Arial", font_size = 10) %>%
  add_header_above(c(" " = 1, "Genetic network" = 3, "Hydrodynamic network" = 3),
                   bold = TRUE) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(0, bold = TRUE)
```

## Table S3 — Concordance classification

```{r ms-table-s3}
ms_concordance <- concordance_classified %>%
  transmute(
    `Region pair`           = region_pair,
    `Genetic connectivity`  = str_to_sentence(genetic_connectivity),
    `Hydrodynamic consensus`= str_to_sentence(hydro_consensus),
    `Agreement`             = agreement
  ) %>%
  arrange(Agreement, `Region pair`)

# Colour-code agreement column
agree_colours <- case_when(
  ms_concordance$Agreement == "Concordant"           ~ "#d4edda",
  ms_concordance$Agreement == "Partially concordant" ~ "#fff3cd",
  ms_concordance$Agreement == "Discordant"           ~ "#f8d7da"
)

kbl(ms_concordance,
    caption = paste0("Table S3. Concordance classification for all overlapping region pairs (n = ",
                     nrow(ms_concordance), "). Jaccard network similarity = ",
                     round(jaccard * 100, 1), "%."),
    align   = c("l","c","c","c")) %>%
  kable_classic(full_width = FALSE, html_font = "Arial", font_size = 10) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(4, background = agree_colours, bold = TRUE)
```

# Session Info

```{r session-info}
sessionInfo()
```
